Missing The Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-11-19 Manuscript.Rmd
Add improved methods in bullet points

---
title: "Missing The Point: Non-Convergence in Iterative Imputation Algorithms"
author: "Hanne Oberman"
date: "7-9-2020"
output:
  bookdown::html_document2
bibliography: ref.bib
---

@@

- the aim is to draw valid inference: to get unbiased, confidence-valid estimates

@@

- the missing data problem is solved by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE

@@

- to obtain valid estimates, both the missing data problem and the scientific problem should be solved correctly

@@

- there is no consensus on the convergence properties of iterative imputation, and there has not been a systematic study on how to evaluate the convergence of these algorithms

- the current practice is visual inspection, but this may not be adequate because 1, 2, 3 (see original text)

- it is difficult because there is not at single point at which convergence is reached, because the algorithm will produce some fluctuation even after convergence (and a ... distribution may not exist because of univariate modeling of multivariate distribution???)

@@

- there are non-convergence identifiers for other iterative algorithms (such as MCMC), but it is not known whether these are appropriate for imputation algorithms as well

@@

- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is not influential, apparently

- since we typically don't have the ground truth (to calculate bias and coverage), we develop guidelines for interpreting the non-convergence identifiers in practice

@@

## Original text

@@

# Identifying non-convergence

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases. [Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences). And add the `ampute` function?]

@@

$$
\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
$$
where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.

@@

We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter that we propose, $\lambda$.

@@

```{r ac, fig.cap = "Autocorrelation (AC) with different parameters.", echo=FALSE}
ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
(ac_means + ac_vars) / (ac_betas + ac_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

@@

Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A). The decrease is substantive until $it \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B). The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $it \geq 10$. We do not observe a systematic difference between missingness conditions here either.

There is more autocorrelation in the scientific estimates than in the chain means and chain variances (see \@ref(fig:ac)C). We observe the highest autocorrelations in conditions where 75% of cases are incomplete. Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$. There is no clear effect of the missingness mechanisms [unnecessary?].

The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $it \geq 20$.

Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
```{r rh, fig.cap = "Potential scale reduction factor (Rh) with different parameters.", echo=FALSE}
rh_means <- simulation_plot(v="r.hat.max.mu.Y", name = "Rh chain means") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_vars <- simulation_plot(v="r.hat.max.sigma.Y", name = "Rh chain variances") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_betas <- simulation_plot(v="r.hat.max.qhat", name = quote("Rh"~beta)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_lambdas <- simulation_plot(v="r.hat.max.lambda", name = quote("Rh"~lambda)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
(rh_means + rh_vars) / (rh_betas + rh_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

@@

We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A). An exception to this observation is the initial increase when $3 \leq it \leq 5$ [interpret?? due to initialization or is there really more mixing initially??]. After the first couple of iterations, the mixing between chain means generally improves until $it \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B). Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $it \geq 30$.

@@

@@

We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A). When $it \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].

Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B). We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].

@@

The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D). Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.

@@

The validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought. We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached. Any further iterations just burn computational resources without improving the statistical inferences.

@@

-   Convergence diagnostics keep improving substantially until it = 20-30

-   Performance measures do not improve after it = 9

@@

- Idea: calculate rhat for each block of 5 it

@@

The scientific estimate reaches the threshold much sooner, when $it=14$ (see \@ref(fig:case)C).

According to the $\widehat{R}$-values of the novel parameter, at least 15 iterations are required (see \@ref(fig:case)D).

# References (incomplete)

>>>>

---
title: "Missing The Point: Non-Convergence in Iterative Imputation Algorithms"
author: "Hanne Oberman"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2
bibliography: ref.bib
---

@@

- the aim is to draw valid inference: to get unbiased, confidence-valid estimates that incorporate the effect of the missingness

@@

- the missing data problem is solved first: by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE

@@

- to obtain valid scientific estimates, both the missing data problem and the scientific problem should be solved correctly

@@

- there is no consensus on the convergence properties of iterative imputation algorithms, and there has not been a systematic study on how to evaluate the convergence of these algorithms

- yet, the inferences rely on algorithmic convergence

- the current practice is visual inspection, but this may be inadequate because of reasons 1, 2, and 3 [see original text]

- it is difficult that there is not at single point at which convergence is reached: the algorithm will produce some fluctuations even after convergence [it's convergence in distribution, not to a point]  <!-- (and a ... distribution may not exist because of univariate modeling of multivariate distribution???) -->

@@

- there are non-convergence identifiers for other iterative algorithms (MCMC), but it is not known whether these are appropriate for imputation algorithms as well

@@

- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is, apparently, not influential

- we develop guidelines for practice to interpret the non-convergence identifiers, because we typically don't have the ground truth (to calculate bias and coverage)

- we show how all this works by means of a case study

@@

## Original text [remove later]

@@

# Identifying non-convergence [remove later]

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios that are defined by the data generating mechanism. 
<!-- : three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??]. -->

@@

The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.

- we use the three missingness mechanisms MCAR, MAR, and MNAR with defaults settings from `mice::ampute()` (e.g., right-tailed MAR)

- we use a multivariate missing data pattern and make 25%, 50%, and 75% of cases incomplete

@@

$$
\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
$$

where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.

@@

- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter

- the parameter can be any statistic that we track across iterations, for example chain means (i.e, the average imputed value per imputation)

- the identifier is a calculation of some sort that quantifies non-convergence

- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains

- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well

- just to be sure, we also use autocorrelation to quantify trending within chains

- we apply these identifiers to four different univariate and multivariate parameters

- the univariate parameters are those commonly used for visual inspection: chain means and chain variances

- one multivariate parameter that is of immediate interest for our estimand is the estimated value itself: regression coefficient beta_1

- we also propose a new multivariate parameter that is not dependent on the model of scientific interest: the first eigenvalue of the variance-covariance matrix of the completed data [explain!]

We use eight different methods to diagnose non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters---chain means, chain variances, $\beta_1$, and $\lambda$.

@@

```{r ac, fig.cap = "Autocorrelation (AC) with different parameters.", echo=FALSE}
ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") + ggplot2::scale_y_continuous(limits = c(0.175, 1))
ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances") + ggplot2::scale_y_continuous(limits = c(0.175, 1))
ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta)) + ggplot2::scale_y_continuous(limits = c(0.175, 1))
ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda)) + ggplot2::scale_y_continuous(limits = c(0.175, 1))
(ac_means + ac_vars) / (ac_betas + ac_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

@@

Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A). The decrease is substantive until $n_{it} \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B). The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$. We do not observe a systematic difference between missingness conditions here either.

There is more autocorrelation in the scientific estimates than in the chain means and chain variances (see \@ref(fig:ac)C). We observe the highest autocorrelations in conditions where 75% of cases are incomplete. Overall, the autocorrelations reach a plateau when $n_{it} \geq 20$ to $30$. There is no clear effect of the missingness mechanisms.

The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $n_{it} \geq 20$.

Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->

```{r rh, fig.cap = "Potential scale reduction factor (Rh) with different parameters.", echo=FALSE}
rh_means <- simulation_plot(v="r.hat.max.mu.Y", name = "Rh chain means") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_vars <- simulation_plot(v="r.hat.max.sigma.Y", name = "Rh chain variances") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_betas <- simulation_plot(v="r.hat.max.qhat", name = quote("Rh"~beta)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
rh_lambdas <- simulation_plot(v="r.hat.max.lambda", name = quote("Rh"~lambda)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
(rh_means + rh_vars) / (rh_betas + rh_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

@@

We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A). An exception to this observation is the initial increase when $3 \leq n_{it} \leq 5$ [interpret?? due to initialization or is there really more mixing initially??]. After the first couple of iterations, the mixing between chain means generally improves until $n_{it} \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B). Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.

@@

In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), bias in the coefficient of determination (panel B), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).

@@

We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A). When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].

Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B). We observe stable estimates in all conditions when $n_{it} \geq 6$ [interpret the over-estimation in MNAR+75% condition?].

@@

The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D). Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.

@@

@@

- Convergence diagnostics keep improving substantially until n_it = 20-30

- Performance measures do not improve after n_it = 9

@@

In short, the validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought. We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached. Any further iterations just burn computational resources without improving the statistical inferences.

@@

The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).

According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).

# References
